% SPDX-License-Identifier: Apache-2.0 OR MIND-UCAL-1.0
% © James Ross Ω FLYING•ROBOTS <https://github.com/flyingrobots>

\chapter*{Colophon}
\addcontentsline{toc}{chapter}{Colophon}

\section*{About This Document}

This \emph{Tour de Code} was generated from the Echo source repository. File
paths and line numbers are accurate as of the document date; code evolves, so
verify against \texttt{HEAD} when in doubt.

\section*{Notation Conventions}

\begin{description}
  \item[Rust paths] Module paths like \texttt{warp\_core::scheduler}
        correspond to \texttt{crates/warp-core/src/scheduler.rs}. Functions like
        \texttt{drain\_for\_tx()} exist inside that file. Submodules like
        \texttt{warp\_core::boaw::shard} map to \texttt{crates/warp-core/src/boaw/shard.rs}.
  \item[Complexity] Big-O notation uses $n$ for input size, $F$ for footprint
        cardinality, $V$ and $E$ for graph vertices and edges.
  \item[Hash functions] ``$H(\cdot)$'' always means BLAKE3 unless otherwise noted.
\end{description}

\section*{Key Invariants to Remember}

\begin{enumerate}
  \item \textbf{Determinism.} Same inputs $\Rightarrow$ same commit hash,
        regardless of worker count or thread scheduling.
  \item \textbf{Content addressing.} Identity derives from content---intents,
        nodes, and commits are all hashed.
  \item \textbf{Canonical ordering.} Radix sort with $(scope\_hash, rule\_id, nonce)$
        keys ensures reproducible execution order.
  \item \textbf{Isolation.} Workers write to private deltas; only the canonical
        merge touches shared state.
\end{enumerate}

\section*{Further Reading}

\begin{itemize}
  \item \textbf{ADR-0007} --- BOAW storage and shard routing
  \item \textbf{Phase 6 planning docs} --- Virtual shards and parallelism model
  \item \textbf{Booklet 02: Core Concepts} --- Higher-level architecture overview
  \item \textbf{Booklet 03: Math Foundations} --- Formal properties and proofs
\end{itemize}

\vfill

\begin{center}
\rule{0.4\linewidth}{0.4pt}\\[6pt]
\textit{Document generated \today.}\\[2pt]
\textit{Commentary by your friendly AI tour guide.}
\end{center}

\clearpage
