% SPDX-License-Identifier: Apache-2.0 OR MIND-UCAL-1.0
% © James Ross Ω FLYING•ROBOTS <https://github.com/flyingrobots>

\chapter{Onboarding Roadmap (First 30–60 Minutes)}

\section{Get the Toolchain}
\begin{itemize}
    \item Install Rust per \texttt{rust-toolchain.toml} (workspace MSRV).
    \item Install Node $\geq$ 18 and \texttt{pnpm} (see \texttt{packageManager} in \texttt{package.json}).
    \item Run \texttt{make hooks} once to enable repo checks.
\end{itemize}

\section{Clone, Smoke Test, and Guard Rails}
\begin{itemize}
    \item \textbf{Clone}: \texttt{git clone} and \texttt{pnpm install} (docs) / \texttt{cargo fetch} (Rust).
    \item \textbf{Docs dev}: \texttt{pnpm docs:dev} to confirm VitePress works.
    \item \textbf{Rust sanity}: \texttt{cargo test -p warp-core -- --list} to ensure toolchain is healthy.
    \item \textbf{Hooks}: verify pre-commit runs (fmt/clippy/docs guard) by committing a no-op change locally.
\end{itemize}

\section{Read This First}
\begin{enumerate}
    \item \textbf{Orientation}: \texttt{architecture-outline.md} (why Echo exists).
    \item \textbf{Determinism contract}: skim \texttt{determinism-invariants.md} for guarantees.
    \item \textbf{Branching model}: \texttt{branch-merge-playbook.md} (merge-only; no rebases/force pushes).
    \item \textbf{Work map}: GitHub Issues and project board for current tasks.
\end{enumerate}

\section{Do One Hands-On Task}
\begin{itemize}
    \item Run a focused test: \texttt{cargo test radix\_drain --package warp-core}.
    \item Explore a doc: open \texttt{docs/public/collision-dpo-tour.html} to see visuals.
\end{itemize}

\section{Next Hour: Choose Your Track}
\begin{itemize}
    \item \textbf{Runtime}: read Scheduler + Storage specs, then run scheduler benchmarks.
    \item \textbf{Docs/Math}: read deterministic math specs; reproduce a figure in TikZ.
    \item \textbf{Ops}: rehearse the merge playbook and run Docs Guard locally.
\end{itemize}
