% SPDX-License-Identifier: Apache-2.0 OR MIND-UCAL-1.0
% © James Ross Ω FLYING•ROBOTS <https://github.com/flyingrobots>
\section{What is Echo?}

Echo is a \textbf{deterministic, multiverse-aware Entity Component System (ECS)} engine. Unlike traditional game engines that prioritize visual fidelity or editor convenience, Echo prioritizes \textbf{determinism}, \textbf{state preservation}, and \textbf{simulation integrity}.

It is built on the concepts described by the \AION{} Foundations Series papers---"Worldline Algebra for Recursive Provenance"  (WARP) graphs---meaning that the entire game state is a WARP graph, and updates are deterministic graph rewrites. This allows for features that are practically impossible in standard engines, such as:

\begin{itemize}
    \item \textbf{Perfect Determinism:} Given the same initial state and inputs, the simulation will evolve identically on any machine (x86, ARM, WASM).
    \item \textbf{Time Travel:} The ability to rollback the simulation to any previous tick, branch time into alternative realities, and merge them back.
    \item \textbf{Renderer Agnosticism:} Echo does not care how it is drawn. It emits an event stream that describes the latest graph writes applied to the \WARP{} State at a given tick.
\end{itemize}

\subsection{Core Philosophy}

Echo was born from earlier prototypes and operates under a set of strict cultural principles:

\begin{enumerate}
    \item \textbf{Tests Are The Spec:} Tests document expected behavior and serve as living documentation---if a test passes, the feature works as intended.
    \item \textbf{Hexagonal Domain Boundary:} Core domain logic is isolated from external interfaces (rendering, networking, persistence) via clean adapter boundaries.
    \item \textbf{We Do Not Rewrite History:} The immutable commit ledger is permanent; every state change is recorded and auditable.
    \item \textbf{Predictable Loop:} Every frame/tick has a deterministic, cryptographically-verifiable state description.
\end{enumerate}

\subsection{Who is Echo For?}

Echo is designed for general use. Its WARP graph foundation makes it particularly suited for:
\begin{itemize}
    \item \textbf{RTS Games:} Where thousands of units must remain in sync across a network.
    \item \textbf{Fighting Games:} Where rollback netcode (GGPO-style) is a requirement, not a feature.
    \item \textbf{Scientific Simulations:} Where reproducibility is critical.
    \item \textbf{Collaborative Tools:} Where multiple users edit a shared state (using the branching/merging capabilities).
\end{itemize}
